---
title: "Example for INTACT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example for INTACT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r deps, include=FALSE}
# Packages used only for figures
plot_pkgs <- c("ggplot2","tidyr","patchwork","ggpubr","tidyselect")
have_plot_pkgs <- vapply(plot_pkgs, requireNamespace, logical(1), quietly = TRUE)
can_plot <- all(have_plot_pkgs)

if (can_plot) {
  # attach for convenience
  for (p in plot_pkgs) library(p, character.only = TRUE)
} else {
  message("Note: skipping plotting sections; missing packages: ",
          paste(plot_pkgs[!have_plot_pkgs], collapse = ", "))
}
```

# Overview

Welcome to **intactPA**!  
This vignette demonstrates how to use the package to harmonize longitudinal
physical activity features collected from multiple sources.

We first show installation, then simulate data with known structure and apply
`INTACT` alongside competing methods, and finally summarize accuracy metrics.

# Installation

If you have not installed the package yet, you can install it from GitHub:

```{r install, eval=FALSE}
# Install from GitHub (uncomment to run)
# install.packages("devtools")
# devtools::install_github("jingru-zhang/intactPA", build_vignettes = TRUE)
```

Load the package:
```{r}
set.seed(12)
library(intactPA)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7.5,
  fig.height = 2.5,
  fig.align = "center"
)
```

# Simulation Example

We simulate subject-level (X part) and visit-level (W part) components, add a
common mean curve and a fixed effect, combine across two sources, and run
`INTACT`. You can optionally introduce missingness by setting `nc <- 1`.

```{r}
nc <- 0                    # set to 1 to introduce missingness
subnum <- c(200, 200)      # subjects per source

# Helper function to generate simulated data
generdata <- function(n, Phi, U, Lam, D, noise = 0) {
  M <- length(U)
  p <- nrow(Phi)
  d <- ncol(Phi)
  data <- data.true <- eij <- vector("list", M)
  
  for (m in 1:M) {
    xi <- matrix(rnorm(n[m] * d), d, n[m])
    data[[m]] <- t(Phi %*% U[[m]] %*% sqrt(Lam * D[[m]]) %*% xi)  # n x p
    eij[[m]] <- matrix(rnorm(n[m] * p), n[m], p) * noise
    data[[m]] <- data[[m]] + eij[[m]]
    data.true[[m]] <- t(Phi %*% sqrt(Lam) %*% xi)
  }
  
  list(data = data, data.true = data.true, eij = eij)
}

# Load example basis functions and parameters:
# Contains: Phixfun, Phiwfun, alpha1, beta1, Ux, Uw, Dx, Dw
data(simuPhifun)

Lamx <- diag(c(3.6, 2.4, 1)) / 2
Lamw <- diag(c(2.0, 1.4, 1.0)) / 2

p <- nrow(Phixfun)
M <- 2
genday <- list(sample(3:7, subnum[1], replace = TRUE),
               sample(3:7, subnum[2], replace = TRUE))
obsnum <- unlist(lapply(genday, sum))

# Generate subject-level and visit-level components
genXdata <- generdata(subnum, Phixfun, Ux, Lamx, Dx, noise = 0.0)
genX <- genXdata$data
genWdata <- generdata(obsnum, Phiwfun, Uw, Lamw, Dw, noise = 0.2)
genW <- genWdata$data

# Assemble full observations per source
genY <- zz.sub <- vector("list", M)
zzall <- c()
cumsub <- c(0, cumsum(subnum))

for (m in 1:M) {
  genY[[m]] <- matrix(0, obsnum[m], p)
  cumday <- c(0, cumsum(genday[[m]]))
  
  for (sub in 1:subnum[m]) {
    subrow <- (cumday[sub] + 1):cumday[sub + 1]
    genY[[m]][subrow, ] <- matrix(genX[[m]][sub, ], genday[[m]][sub], p, byrow = TRUE) +
                           genW[[m]][subrow, ]
  }
  
  # Subject-level covariate and fixed effect
  zz.sub[[m]] <- rnorm(subnum[m], mean = 1, sd = 0.5)
  zz <- rep(zz.sub[[m]], times = genday[[m]])
  genY[[m]] <- genY[[m]] + zz %*% t(beta1) + matrix(alpha1, obsnum[m], p, byrow = TRUE)
  
  rownames(genY[[m]]) <- rep((cumsub[m] + 1):cumsub[m + 1], times = genday[[m]])
  zzall <- c(zzall, zz)
}

# Package into data.frame and feature matrix
mydf0 <- data.frame(
  subid = c(rownames(genY[[1]]), rownames(genY[[2]])),
  zmat  = zzall,
  batch = c(rep(0, nrow(genY[[1]])), rep(1, nrow(genY[[2]])))
)

Yraw <- rbind(genY[[1]], genY[[2]])
rownames(Yraw) <- mydf0$subid

# Optionally introduce missingness
if (nc == 1) {
  Yraw.nc <- Yraw
  Yraw.nc[sample(length(Yraw.nc), 0.3 * nrow(Yraw.nc) * ncol(Yraw.nc))] <- NA
  Yraw <- Yraw.nc
}

# INTACT configuration
formula      <- "zmat"
needlog      <- 0
comparisons  <- 1  # also compute competing methods
act          <- 1
wpve         <- 0.5
dd           <- NULL

# Run INTACT
reall <- INTACT(mydf0, Yraw, formula, needlog, comparisons, act, wpve, dd)

# Extract selected results
multipar    <- reall$reintact$multipar
Y_intact    <- reall$reintact$Y
Eta_intact  <- reall$reintact$Eta
Y_intact0   <- reall$reintact0$Y
Eta_intact0 <- reall$reintact0$Eta

# Competing methods
Y_raw     <- reall$reraw$Y
Eta_raw   <- reall$reraw$Eta
Y_sim     <- reall$resim$Y
Eta_sim   <- reall$resim$Eta
Y_quan    <- reall$requan$Y
Eta_quan  <- reall$requan$Eta
Y_lcombat <- reall$relcombat$Y
Eta_lcombat <- reall$relcombat$Eta
```

# Accuracy Metrics

We can now evaluate the estimation accuracy of the harmonized data.  

The function `getMVDV` compares estimated covariance structures with the truth 
and across multiple sources.

```{r}
Sim_raw     <- getMVDV(Phixfun, Lamx, Phiwfun, Lamw, Eta_raw,     multipar)
Sim_sim     <- getMVDV(Phixfun, Lamx, Phiwfun, Lamw, Eta_sim,     multipar)
Sim_quan    <- getMVDV(Phixfun, Lamx, Phiwfun, Lamw, Eta_quan,    multipar)
Sim_lcombat <- getMVDV(Phixfun, Lamx, Phiwfun, Lamw, Eta_lcombat, multipar)
Sim_intact  <- getMVDV(Phixfun, Lamx, Phiwfun, Lamw, Eta_intact,  multipar)
Sim_intact0 <- getMVDV(Phixfun, Lamx, Phiwfun, Lamw, Eta_intact0, multipar)

# Print first four entries (MV.X, MV.W, DV.X, DV.W)
unlist(Sim_raw[1:4])
unlist(Sim_sim[1:4])
unlist(Sim_quan[1:4])
unlist(Sim_lcombat[1:4])
unlist(Sim_intact[1:4])
unlist(Sim_intact0[1:4])
```

# Visualizations
## Mean curves across methods
```{r,eval=can_plot}
# plotting helper
plotmcurve <- function(usephi, tx, ss = 9) {
  df <- as.data.frame(usephi)
  colnames(df) <- c("curve1", "curve2")
  df$t <- 1:nrow(df)

  df_long <- tidyr::pivot_longer(
    df,
    cols = tidyselect::starts_with("curve"),
    names_to = "curve",
    values_to = "value"
  )

  ggplot(df_long, aes(x = t, y = value, color = curve, linetype = curve)) +
    geom_line() +
    scale_color_manual(values = c("curve1" = "black", "curve2" = "blue")) +
    scale_linetype_manual(values = c("curve1" = "dashed", "curve2" = "dotted")) +
    labs(title = tx, x = "", y = "") +
    theme_minimal() +
    theme(
      panel.background = element_blank(),
      plot.background = element_blank(),
      panel.grid = element_blank(),
      panel.border = element_rect(fill = NA, color = "black"),
      axis.line = element_blank(),
      axis.ticks = element_line(),
      axis.text.x = element_text(size = ss),
      plot.title = element_text(hjust = 0.5, size = 10),
      legend.position = "none"
    )
}

trim <- function(p, margin_pt = 0) {
  p + theme(plot.margin = unit(rep(margin_pt, 4), "pt"))
}

ss = 6.3
mycurve = exp(cbind(colMeans(Y_raw[[1]]),colMeans(Y_raw[[2]])))-1
mcurve.raw = plotmcurve(mycurve,'raw',ss)
mycurve = exp(cbind(colMeans(Y_sim[[1]]),colMeans(Y_sim[[2]])))-1
mcurve.sim = plotmcurve(mycurve,'standard',ss)
mycurve = exp(cbind(colMeans(Y_quan[[1]]),colMeans(Y_quan[[2]])))-1
mcurve.quan = plotmcurve(mycurve,'quantile',ss)
mycurve = exp(cbind(colMeans(Y_lcombat[[1]]),colMeans(Y_lcombat[[2]])))-1
mcurve.com = plotmcurve(mycurve,'L-ComBat',ss)
mycurve = exp(cbind(colMeans(Y_intact[[1]]),colMeans(Y_intact[[2]])))-1
mcurve.int = plotmcurve(mycurve,'INTACT',ss)
mycurve = exp(cbind(colMeans(Y_intact0[[1]]),colMeans(Y_intact0[[2]])))-1
mcurve.int0 = plotmcurve(mycurve,expression(INTACT[0]),ss)

mcurveall <- (
  trim(mcurve.raw)  +
  trim(mcurve.sim)  +
  trim(mcurve.quan) +
  trim(mcurve.com)  +
  trim(mcurve.int)  +
  trim(mcurve.int0)
) + plot_layout(ncol = 6, guides = "collect")

mcurveall
```

## Covariance difference heatmaps
```{r,eval=can_plot}
getVardiff = function(Y){
  M = 2
  var.l1 = var.l2 = vector("list",M)
  for(m in 1:M){
    id = rownames(Y[[m]])
    fpca <- fastmfpca(Y = as.matrix(Y[[m]]), id = id, twoway = FALSE, pve = 1)
    funbasis = as.matrix(fpca$funbasis)
    var.l1[[m]] = funbasis %*% fpca$Thetamat[[1]] %*% t(funbasis)
    var.l2[[m]] = funbasis %*% fpca$Thetamat[[2]] %*% t(funbasis)
  }
  dvar.l1 = var.l1[[1]] - var.l1[[2]]
  dvar.l2 = var.l2[[1]] - var.l2[[2]]
  diffvar = list(dvar.l1, dvar.l2, var.l1, var.l2)
  names(diffvar) = c("l1: M1 vs M2","l2: M1 vs M2","var.l1","var.l2")
  return(diffvar)
}

diff_raw = getVardiff(Eta_raw)
diff_sim = getVardiff(Eta_sim)
diff_quan = getVardiff(Eta_quan)
diff_lcombat = getVardiff(Eta_lcombat)
diff_intact = getVardiff(Eta_intact)
diff_intact0 = getVardiff(Eta_intact0)

myheat = vector("list",2)
yy = c("X part","W part")
for (chl in 1:2) {
  t0 = as.matrix(diff_raw[[chl]])*multipar^2
  t1 = as.matrix(diff_sim[[chl]])*multipar^2
  t2 = as.matrix(diff_quan[[chl]])*multipar^2
  t3 = as.matrix(diff_lcombat[[chl]])*multipar^2
  t4 = as.matrix(diff_intact[[chl]])*multipar^2
  t5 = as.matrix(diff_intact0[[chl]])*multipar^2

  num.method = 6
  dimp = nrow(t0)
  value = c(as.vector(t0),as.vector(t1),as.vector(t2),
            as.vector(t3),as.vector(t4),as.vector(t5)) 
  method = factor(rep(c("raw","standard","quantile",
                        "L-ComBat","INTACT","INTACT0"),each=dimp^2),
                  levels = c("raw","standard","quantile",
                             "L-ComBat","INTACT","INTACT0"))

  IDx = rep(rep(1:dimp,dimp),num.method)
  IDy = rep(rep(1:dimp,each=dimp),num.method) 
  dataf = data.frame(IDx,IDy,value,method)
  levels(dataf$method) <- c("raw","standard","quantile",
                            "L-ComBat","INTACT","INTACT[0]")

  myheat[[chl]] = ggplot(dataf, aes(IDx,IDy)) + 
    facet_grid(.~method)  +
    geom_tile(aes(fill = value)) + theme_bw() + theme(panel.grid = element_blank(), axis.line = element_blank()) +
    theme(strip.text = element_text(colour = 'black', size = rel(1.2)), strip.background = element_blank()) +
    theme(axis.text.x = element_blank(), axis.ticks.x=element_blank()) + xlab("") +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + ylab(yy[chl]) + 
    theme(legend.title = element_blank()) +
    scale_fill_gradient2(low = '#0571b0', high = '#ca0020',midpoint=0) +
    facet_grid(.~method, labeller = label_parsed)
}

myheat[[1]] = myheat[[1]] + theme(axis.text=element_text(size=10),
                                  axis.title=element_text(size=12))+
  theme(plot.title = element_text(hjust = 0.5,size=12))+
  theme(legend.text=element_text(size=10))+
  theme(strip.text.x = element_text(size = 12))
myheat[[2]] = myheat[[2]] + theme(axis.text=element_text(size=10),
                                  axis.title=element_text(size=12))+
  theme(plot.title = element_text(hjust = 0.5,size=12))+
  theme(legend.text=element_text(size=10))+
  theme(strip.text.x = element_text(size = 12))

common_limits <- c(-0.03, 0.03)
common_scale <- scale_fill_gradient2(
  low       = "#0571b0",
  mid       = "white",
  high      = "#ca0020",
  midpoint  = 0,
  limits    = common_limits,
  name      = NULL
)
myheat[[1]] <- myheat[[1]] + common_scale
myheat[[2]] <- myheat[[2]] + common_scale

# Heatmaps of covariance differences between sources
# the first row is for X part and the second row is for W part
heatdiffall = myheat[[1]] + myheat[[2]] +
  plot_layout(ncol = 1, nrow = 2, guides="collect") & 
  theme(legend.position = "right")
heatdiffall 
```

## Estimation results
```{r,eval=can_plot}
# d is 2-dimensional vector giving the estimated numbers of components for the X- and W-parts, respectively.
d = reall$d
d

fixeffs = reall$fixeffs
alpha1.hat = fixeffs[1,]*multipar
beta1.hat  = fixeffs[2,]*multipar

# Common mean curve (true vs estimated)
ss = 6.3
alphacurve = plotmcurve(cbind(alpha1, alpha1.hat), expression({alpha}(t)), ss)
betacurve  = plotmcurve(cbind(beta1,  beta1.hat ), expression({beta}(t)),  ss)

alphacurve
betacurve
```

## Common eigenfunctions (true vs estimated)
```{r,eval=can_plot}
Xi.x = -reall$Xi.x
Xi.w = -reall$Xi.w

Phixfun.a = Phiwfun.a = vector("list",2)
for (i in 1:2) {
  Phixfun.a[[i]] = Phixfun %*% Ux[[i]]
  Phiwfun.a[[i]] = Phiwfun %*% Uw[[i]]
}

plotphi <- function(usephi, tx, ss = 9) {
  df <- as.data.frame(usephi)
  colnames(df) <- c("phicom", "phi1", "phi2", "phicomest")
  df$t <- 1:nrow(df)

  df_long <- tidyr::pivot_longer(
    df,
    cols = tidyselect::starts_with("phi"),
    names_to = "curve",
    values_to = "value"
  )

  ggplot(df_long, aes(x = t, y = value, color = curve, linetype = curve)) +
    geom_line() +
    scale_color_manual(values = c("phicom" = "red", "phi1" = "black",
                                  "phi2" = "blue", "phicomest" = "purple")) +
    scale_linetype_manual(values = c("phicom" = "solid", "phi1" = "dashed",
                                     "phi2" = "dotted", "phicomest" = "solid")) +
    labs(title = tx, x = "", y = "") +
    theme_minimal() +
    theme(
      panel.background = element_blank(),
      plot.background = element_blank(),
      panel.grid = element_blank(),
      panel.border = element_rect(fill = NA, color = "black"),
      axis.line = element_blank(),
      axis.ticks = element_line(),
      axis.text.x = element_text(size = ss),
      plot.title = element_text(hjust = 0.5, size = 10),
      legend.position = "none"
    )
}

gphi.x = vector("list",3)
for(j in 1:3){
  usephi = cbind(Phixfun[,j],Phixfun.a[[1]][,j],Phixfun.a[[2]][,j],Xi.x[,j])
  tx = paste('Eigenfunction ',j,' for X part',sep="")
  gphi.x[[j]] = plotphi(usephi,tx)
}
gphi.w = vector("list",3)
for(j in 1:3){
  usephi = cbind(Phiwfun[,j],Phiwfun.a[[1]][,j],Phiwfun.a[[2]][,j],Xi.w[,j])
  tx = paste('Eigenfunction ',j,' for W part',sep="")
  gphi.w[[j]] = plotphi(usephi,tx)
}

phix.all = gphi.x[[1]] + gphi.x[[2]] + gphi.x[[3]] + plot_layout(ncol = 3,nrow = 1,guides = "collect")
phiw.all = gphi.w[[1]] + gphi.w[[2]] + gphi.w[[3]] + plot_layout(ncol = 3,nrow = 1,guides = "collect")

ggarrange(phix.all, phiw.all, ncol = 1, labels = c(""), font.label = list(size = 20))
```












